---
title: "Missing Data"
format: 
   html:
     df-print: default
---

# Preamble

```{r}
#| include: false

options(scipen = 999) # suppress scientific notation
```

## Load Libraries

```{r}
library("mice")
library("miceadds")
library("micemd")
library("lme4")
library("lmerTest")
library("broom.mixed")
library("lavaan")
library("lavaan.mi")
library("psych")
library("parallelly")
library("tidyverse")
```

## Load Data

The data for this example were [simulated](simulate-data.qmd):

```{r}
mydata_long <- read.csv("./data/mydata_long.csv")
```

# Full Information Maximum Likelihood {#sec-fiml}

For examples using full information maximum likelihood (FIML), see examples of structural equation models [here](structural-equation-models.qmd).
You can include cases with missingness on predictors by specifying `fixed.x = FALSE`.

# Multiple Imputation {#sec-multipleImputation}

## Describe Missing Data

```{r}
#| eval: false

psych::describe(mydata_long)
```

```{r}
#| echo: false

rmarkdown::paged_table(psych::describe(mydata_long))
```

```{r}
mice::md.pattern(mydata_long, rotate.names = TRUE)
```

## Pre-Imputation Setup

### Specify Variables to Impute

```{r}
varsToImpute <- c("ses","sleep","depression","anxiety")
Y <- varsToImpute
```

### Specify Number of Imputations

```{r}
numImputations <- 5 # generally use 100 imputations; this example uses 5 for speed
```

### Detect Cores

```{r}
#| include: false

num_cores <- parallelly::availableCores()
num_true_cores <- parallelly::availableCores(logical = FALSE)
```

```{r}
#| eval: false

num_cores <- parallelly::availableCores() - 1
num_true_cores <- parallelly::availableCores(logical = FALSE) - 1
```

### Specifying the Imputation Method

You can specify any of various methods to impute each variable.
In this case, we use a multilevel imputation (to account for longitudinal data) form of predictive mean matching (`2l.pmm` from the `miceadds` package).

```{r}
imputationMethods <- mice::make.method(mydata_long)
imputationMethods[1:length(imputationMethods)] <- ""
imputationMethods[Y] <- "2l.pmm" # specify the imputation method here; this can differ by outcome variable
imputationMethods
```

### Specifying the Predictor Matrix

A predictor matrix is a matrix of values, where:

- columns with non-zero values are predictors of the variable specified in the given row
- the diagonal of the predictor matrix should be zero because a variable cannot predict itself

The values are:

- NOT a predictor of the outcome: `0`
- cluster variable: `-2`
- fixed effect of predictor: `1`
- fixed effect and random effect of predictor: `2`
- include cluster mean of predictor in addition to fixed effect of predictor: `3`
- include cluster mean of predictor in addition to fixed effect and random effect of predictor: `4`

```{r}
predictorMatrix <- mice::make.predictorMatrix(mydata_long)
predictorMatrix[1:nrow(predictorMatrix), 1:ncol(predictorMatrix)] <- 0 # make everything zero by default to manually specify all predictors used for imputation
predictorMatrix[Y, "ID"] <- (-2) # cluster variable
predictorMatrix[Y, "age"] <- 2 # random effect predictor
predictorMatrix[Y, "sexMale"] <- 1 # fixed effect predictor
predictorMatrix[Y, Y] <- 1 # fixed effect predictor; make sure all predictors predict each other

diag(predictorMatrix) <- 0 # don't have a variable predicting itself
predictorMatrix
```

## Run Imputation

### Sequential Processing

```{r}
mi_mice <- mice::mice(
  as.data.frame(mydata_long),
  method = imputationMethods,
  predictorMatrix = predictorMatrix,
  m = numImputations,
  maxit = 5, # generally use 100 maximum iterations; this example uses 5 for speed
  seed = 52242)
```

### Parallel Processing

```{r}
mi_parallel_mice <- mice::futuremice(
  as.data.frame(mydata_long),
  method = imputationMethods,
  predictorMatrix = predictorMatrix,
  m = numImputations,
  maxit = 5, # generally use 100 maximum iterations; this example uses 5 for speed
  parallelseed = 52242,
  n.core = num_cores,
  packages = "miceadds")
```

## Imputation Diagnostics

### Logged Events

```{r}
mi_mice$loggedEvents
```

### Trace Plots

On convergence, the streams of the trace plots should intermingle and be free of any trend (at the later iterations).
Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence.

<https://stefvanbuuren.name/fimd/sec-algoptions.html> (archived at <https://perma.cc/4S54-465R>)

```{r}
plot(mi_mice)
```

### Density Plots

```{r}
mice::densityplot(mi_mice)
```

### Strip Plots

```{r}
mice::stripplot(mi_mice)

mice::stripplot(mi_mice, depression ~ .imp)
```

## Post-Processing

### Modify/Create New Variables

```{r}
mi_mice_long <- mice::complete( # return the completed (imputed) data
  mi_mice,
  action = "long",
  include = TRUE)

mi_mice_long <- mi_mice_long |> # long form for mixed-effects modeling
  group_by(ID, .imp) |>
  mutate(
    sexFactor = factor(sex),
    ageCentered = age - min(age, na.rm = TRUE),
    ageCenteredSquared = ageCentered ^ 2,
    ses_personMean = mean(ses, na.rm = TRUE),
    ses_personMeanCentered = ses - ses_personMean,
    sleep_personMean = mean(sleep, na.rm = TRUE),
    sleep_personMeanCentered = sleep - sleep_personMean,
    depression_personMean = mean(depression, na.rm = TRUE),
    depression_personMeanCentered = depression - depression_personMean,
    anxiety_personMean = mean(anxiety, na.rm = TRUE),
    anxiety_personMeanCentered = anxiety - anxiety_personMean,
  ) |>
  ungroup()

mi_mice_wide <- mi_mice_long |> # wide form for structural equation modeling: widen by time
  select(-.id, -age, -ageCentered, -ageCenteredSquared) |> # drop time-varying variables
  tidyr::pivot_wider(
    names_from  = timepoint,
    values_from = c(ses, ses_personMeanCentered, sleep, sleep_personMeanCentered, depression, depression_personMeanCentered, anxiety, anxiety_personMeanCentered)
  )
```

### Convert to `mids` object

```{r}
mi_mice_long_mids <- mice::as.mids(mi_mice_long)

mi_mice_wide_mids <- mice::as.mids(mi_mice_wide)
```

## Fit Models to Multiple Imputed Data

### Mixed-Effects Models

#### Fit Model

```{r}
linearGCM_mice <- with(
  data = mi_mice_long_mids,
  expr = lmerTest::lmer(
    formula = depression ~ sexFactor + ageCentered + sexFactor:ageCentered + 
      sleep + sleep:ageCentered + sexFactor:sleep + sexFactor:sleep:ageCentered + 
      (1 + ageCentered | ID), # random intercepts and slopes; sex as a fixed-effect predictor of the intercepts and slopes
  )
)
```

#### Model Summary

```{r}
linearGCM_mice
```

| Column label | Definition                                                          |
|:-------------|:--------------------------------------------------------------------|
| `m`          | Number of imputations                                               |
| `estimate`   | Pooled complete data estimate                                       |
| `ubar`       | Within-imputation variance of `estimate`                            |
| `b`          | Between-imputation variance of `estimate`                           |
| `t`          | Total variance, of `estimate`                                       |
| `dfcom`      | Degrees of freedom in complete data                                 |
| `df`         | Residual degrees of freedom for hypothesis testing of $t$-statistic |
| `riv`        | Relative increase in variance due to nonresponse                    |
| `lambda`     | Proportion of total variance due to missingness                     |
| `fmi`        | Fraction of missing information                                     |

: Elements of pooled object {#tbl-pooledObjects}

```{r}
linearGCM_mice_pooled <- mice::pool(linearGCM_mice)

linearGCM_mice_pooled
```

```{r}
#| eval: false

summary(linearGCM_mice_pooled)
```

```{r}
#| echo: false

rmarkdown::paged_table(summary(linearGCM_mice_pooled))
```

### Structural Equation Models

#### Model Syntax

```{r}
linearGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  slope =~ 0*depression_1 + 1*depression_2 + 2*depression_3 + 3*depression_4

  # Regression paths
  intercept ~ sexMale
  slope ~ sexMale
  
  # Time-varying covariates
  depression_1 ~ sleep_1
  depression_2 ~ sleep_2
  depression_3 ~ sleep_3
  depression_4 ~ sleep_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0
  depression_2 ~ 0
  depression_3 ~ 0
  depression_4 ~ 0
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  slope ~ 1
'
```

#### Fit Model

```{r}
linearGCM_fit_mice <- lavaan.mi::sem.mi(
  linearGCM_syntax,
  data = mi_mice_wide_mids,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE # estimate means/intercepts
)
```

#### Model Summary

```{r}
summary(
  linearGCM_fit_mice,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```
