---
title: "Mixed-Effects Models"
---

For an overview of mixed modeling, see my chapter on "[Mixed Modeling](https://isaactpetersen.github.io/Fantasy-Football-Analytics-Textbook/mixed-models.html)" in the following textbook that is available for free online:

Petersen, I. T. (2025). *Fantasy football analytics: Statistics, prediction, and empiricism using R*. University of Iowa Libraries. <https://isaactpetersen.github.io/Fantasy-Football-Analytics-Textbook>

# Preamble

```{r}
#| include: false

options(scipen = 999) # suppress scientific notation
```

## Load Libraries

```{r}
library("lme4")
library("lmerTest")
library("mgcv")
library("easystats")
library("parallelly")
library("MuMIn")
library("tidyverse")
```

## Load Data

The data for this example were [simulated](simulate-data.qmd):

```{r}
mydata_long <- read.csv("./data/mydata_long.csv")
```

# Research Questions {#sec-researchQuestions}

1. What is the form of adolescents' depressive symptom trajectories across ages 14 to 17?
1. Do the trajectories differ between boys and girls?
1. Does how much sleep the adolescent receives predict their depressive symptoms over time?
1. Does the association between sleep and depressive symptoms differ between boys and girls?
1. What is the association between sleep and depressive symptoms at the within-individual level versus between-individual level?

# Pre-Model Computation {#sec-premodelComputation}

It can be helpful to center the age/time variable so that the intercept in a growth curve model has meaning.
For instance, we can subtract the youngest participant age to set the intercepts to be the earliest age in the sample.

```{r}
mydata_long$sexFactor <- factor(mydata_long$sex) # create a factor variable for sex

mydata_long$ageCentered <- mydata_long$age - min(mydata_long$age, na.rm = TRUE) # shift age so that intercepts represent the earliest age in the sample

mydata_long$ageCenteredSquared <- mydata_long$ageCentered ^ 2
```

# Linear Growth Curve Model {#sec-linearGCM}

## Fit the Model

In estimating linear growth curves of depression symptoms, this model fits:

1. random intercepts and slopes for subjects,
1. `sex` as a time-invariant fixed-effect predictor of the intercepts and slopes, and
1. `sleep` as a time-varying fixed-effect predictor of depression, with effects allowed to vary over age and by sex, including two‐ and three‐way interactions with age and sex.

```{r}
linearGCM <- lmerTest::lmer(
  depression ~ sexFactor + ageCentered + sexFactor:ageCentered + 
    sleep + sleep:ageCentered + sexFactor:sleep + sexFactor:sleep:ageCentered + 
    (1 + ageCentered | ID),
  data = mydata_long,
  na.action = na.exclude)
```

## Model Summary

```{r}
summary(linearGCM)

print(effectsize::standardize_parameters(
  linearGCM,
  method = "refit"),
  digits = 2)

performance::r2(linearGCM)
```

## Plots

```{r}
newData <- expand.grid(
  sexFactor = c("female", "male"),
  age = c(
    min(mydata_long$age, na.rm = TRUE),
    max(mydata_long$age, na.rm = TRUE)),
  sleepFactor = c("low sleep", "average sleep", "high sleep")
)

newData$ageCentered <- newData$age - min(newData$age)

newData$sleep <- NA
newData$sleep[which(newData$sexFactor == "female" & newData$age == 14 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 14 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age == 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age == 14 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 14 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age == 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age == 14 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 14 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age == 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age == 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$modelImpliedDepression <- predict( # predict.merMod
  linearGCM,
  newdata = newData,
  re.form = NA # include no random effects
)
```

### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor)) +
  geom_line() + # fit lines to the model-implied points
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

mydata_long$modelImpliedDepression <- predict( # predict.merMod
  linearGCM,
  newdata = mydata_long,
  re.form = NULL # include all random effects
)

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID,
    color = sexFactor)) +
  geom_smooth( # fit smooth to each participant's points
    method = "lm",
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID)) +
  geom_smooth( # fit smooth to each participant's points
    method = "lm",
    se = FALSE,
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_line( # fit lines to the model-implied points
    data = newData |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = modelImpliedDepression,
      group = sexFactor,
      color = sexFactor),
    linewidth = 2) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor,
    linetype = sleepFactor)) +
  geom_line() + # fit lines to the model-implied points
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  theme_classic()
```

# Polynomial (Quadratic) Growth Curve Model {#sec-quadratic}

## Fit the Model

In estimating polynomial (in this case, quadratic) growth curves of depression symptoms, this model fits:

1. random intercepts and random linear slopes and fixed quadratic slopes for subjects,
1. `sex` as a time-invariant fixed-effect predictor of the intercepts and slopes, and
1. `sleep` as a time-varying fixed-effect predictor of depression, with effects allowed to vary over age and by sex, including two‐ and three‐way interactions with age and sex.

```{r}
quadraticGCM <- lmerTest::lmer(
  depression ~ sexFactor + ageCentered + ageCenteredSquared + 
    sexFactor:ageCentered + sexFactor:ageCenteredSquared + sleep + 
    sleep:ageCentered + sleep:ageCenteredSquared + sexFactor:sleep + 
    sexFactor:sleep:ageCentered + sexFactor:sleep:ageCenteredSquared + 
    (1 + ageCentered | ID), # model failed to converge with random quadratic slopes (ageCenteredSquared)
  data = mydata_long,
  na.action = na.exclude)
```

## Model Summary

```{r}
summary(quadraticGCM)

print(effectsize::standardize_parameters(
  quadraticGCM,
  method = "refit"),
  digits = 2)

performance::r2(quadraticGCM)
```

## Plots

```{r}
newData <- expand.grid(
  sexFactor = c("female", "male"),
  age = 14:17,
  sleepFactor = c("low sleep", "average sleep", "high sleep")
)

newData$ageCentered <- newData$age - min(newData$age)
newData$ageCenteredSquared <- newData$ageCentered ^ 2

newData$sleep <- NA
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$modelImpliedDepression <- predict( # predict.merMod
  quadraticGCM,
  newdata = newData,
  re.form = NA # include no random effects
)
```

### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor)) +
  geom_smooth( # fit quadratic smooth to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

mydata_long$modelImpliedDepression <- predict( # predict.merMod
  quadraticGCM,
  newdata = mydata_long,
  re.form = NULL # include all random effects
)

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID,
    color = sexFactor)) +
  geom_smooth( # fit quadratic smooth to each participant's points
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID)) +
  geom_smooth( # fit quadratic smooth to each participant's points
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_smooth( # fit quadratic smooth to the model-implied points; use geom_line() to fit a non-smooth line
    data = newData |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = modelImpliedDepression,
      group = sexFactor,
      color = sexFactor),
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 2) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor,
    linetype = sleepFactor)) +
  geom_smooth( # fit quadratic smooth to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  guides(
    linetype = guide_legend(
      override.aes = list(color = "black")
    )
  ) +
  theme_classic()
```

# Generalized Additive Mixed Model {#sec-gamm}

Generalized additive mixed models (GAMMs) can be estimated using the `mgcv::gamm()` and `gamm4::gamm4()` functions.

## Fit the Model

```{r}
#| include: false

num_cores <- parallelly::availableCores()
num_true_cores <- parallelly::availableCores(logical = FALSE)
```

```{r}
#| eval: false

num_cores <- parallelly::availableCores() - 1
num_true_cores <- parallelly::availableCores(logical = FALSE) - 1
```

In estimating generalized additive growth models of depression symptoms, this model includes:

1. random intercepts and random linear and quadratic slopes for subjects,
1. sex‐specific smooth functions of age, `sex` and `sleep` as fixed effects with their interaction, and
1. sex‐specific smooth interactions between age and sleep, along with subject‐specific smooth deviations over age.

```{r}
gamGCM <- mgcv::gamm(
  depression ~ sexFactor + sleep + sexFactor:sleep +
    
    # sex-specific trajectories
    s(ageCentered, by = sexFactor, k = 4) +
    
    # sex-specific effects of sleep on trajectories
    ti(ageCentered, sleep, by = sexFactor, k = 4),
    
    # random effects
    random = list(ID = ~ 1 + ageCentered),
  data = mydata_long,
  nthreads = num_cores)

gamGCM_standardized <- mgcv::gamm(
  scale(depression) ~ sexFactor + scale(sleep) + sexFactor:scale(sleep) +
    
    # sex-specific trajectories
    s(scale(ageCentered), by = sexFactor, k = 4) +
    
    # sex-specific effects of sleep on trajectories
    ti(scale(ageCentered), scale(sleep), by = sexFactor, k = 4),
    
    # random effects
    random = list(ID = ~ 1 + scale(ageCentered)),
  data = mydata_long,
  nthreads = num_cores)
```

## Model Summary

```{r}
gamGCM

gamGCM_standardized

performance::r2(gamGCM)
```

## Plots

```{r}
newData <- expand.grid(
  sexFactor = c("female", "male"),
  age = seq(
    from = min(mydata_long$age, na.rm = TRUE),
    to = max(mydata_long$age, na.rm = TRUE),
    length.out = 10000
  ),
  sleepFactor = c("low sleep", "average sleep", "high sleep")
)

newData$ageCentered <- newData$age - min(newData$age)

newData$sleep <- NA
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "average sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "low sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) - sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

newData$sleep[which(newData$sexFactor == "female" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 14 & newData$age < 15 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 14)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 15 & newData$age < 16 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 15)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 16 & newData$age < 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 16)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "female" & newData$age >= 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "female" & mydata_long$age == 17)], na.rm = TRUE)
newData$sleep[which(newData$sexFactor == "male" & newData$age >= 17 & newData$sleepFactor == "high sleep")] <- mean(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE) + sd(mydata_long$sleep[which(mydata_long$sexFactor == "male" & mydata_long$age == 17)], na.rm = TRUE)

# add a dummy id
newData$ID <- mydata_long$ID[1]

newData$modelImpliedDepression <- predict( # predict.gam
  gamGCM,
  newdata = newData
)
```

### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor)) +
  geom_smooth( # fit GAM smooth to the model-implied points; use geom_line() to fit a non-smooth line
    method = "gam",
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

mydata_long$modelImpliedDepression <- predict( # predict.gam
  gamGCM,
  newdata = mydata_long
)

gamMethod <- function(formula, data, weights, ...) {
  if(nrow(data) >= 4){ # gam smooth if 4+ timepoints
    mgcv::gam(y ~ s(x, bs = "cs", k = 3), data = data, ...)
  } else if(nrow(data) == 3){  # quadratic smooth if 3 timepoints
    lm(y ~ x + I(x^2), data = data, ...)
  } else{ # linear if fewer than 3 timepoints
    lm(y ~ x, data = data, ...)
  }
}

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID,
    color = sexFactor)) +
  geom_smooth( # fit GAM smooth to each participant's points
    method = gamMethod,
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = mydata_long,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    group = ID)) +
  geom_smooth( # fit GAM smooth to each participant's points
    method = gamMethod,
    se = FALSE,
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_smooth( # fit GAM smooth to the model-implied points; use geom_line() to fit a non-smooth line
    data = newData |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = modelImpliedDepression,
      group = sexFactor,
      color = sexFactor),
    method = "gam",
    linewidth = 2) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor,
    linetype = sleepFactor)) +
  geom_smooth( # fit GAM smooth to the model-implied points; use geom_line() to fit a non-smooth line
    method = "gam",
    se = FALSE) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  guides(
    linetype = guide_legend(
      override.aes = list(color = "black")
    )
  ) +
  theme_classic()
```

# Compare Models {#sec-compareModels}

```{r}
anova(
  linearGCM,
  quadraticGCM
)

AIC(
  linearGCM,
  quadraticGCM,
  gamGCM
)

MuMIn::AICc(
  linearGCM,
  quadraticGCM,
  gamGCM
)
```

# Disaggregation of Within-Person and Between-Person Effects {#sec-disaggregatingWithinBetween}

## Pre-Model Computations

```{r}
mydata_long <- mydata_long |>
  group_by(ID) |>
  mutate(
    sleep_personMean = mean(sleep, na.rm = TRUE),
    sleep_personMeanCentered = sleep - sleep_personMean
  ) |>
  ungroup()
```

## Fit the Model

In disaggregating within- and between-person effects, this model fits:

1. random intercepts and random linear slopes for subjects,
1. `sex` and `sleep_personMean` as time-invariant fixed-effect predictors of the intercepts and slopes, and
1. `sleep_personMeanCentered` as a time-varying fixed-effect predictor of depression.

```{r}
withinBetweenPersonSleepModel_basic <- lmerTest::lmer(
  depression ~ sleep_personMean + sexFactor +              # time-invariant predictors of the intercepts
    sleep_personMeanCentered + ageCentered +               # time-varying predictors
    sleep_personMean:ageCentered + sexFactor:ageCentered + # time-invariant predictors of the slopes
    (1 + ageCentered | ID),
  data = mydata_long,
  na.action = na.exclude)
```

However, based on our modeling in @sec-quadratic and our model comparison in @sec-compareModels, we determined that a quadratic model fit significantly better than a linear model, so we can extend the model to a quadratic model.

In disaggregating within- and between-person effects, this model fits:

1. random intercepts and random linear slopes and fixed quadratic slopes for subjects,
1. `sex` and `sleep_personMean` (and their interaction) as time-invariant fixed-effect predictors of the intercepts and slopes, and
1. `sleep_personMeanCentered` as a time-varying fixed-effect predictor of depression, with effects allowed to vary over age and by sex, including two‐ and three‐way interactions with age and sex.

The `*` symbol (unlike the `:` symbol) estimates all lower-order main effects and interactions, to simplify the syntax.

```{r}
withinBetweenPersonSleepModel <- lmerTest::lmer(
  depression ~ sexFactor*sleep_personMean*ageCentered + 
    sexFactor*sleep_personMean*ageCenteredSquared + 
    sexFactor*sleep_personMeanCentered*ageCentered + 
    sexFactor*sleep_personMeanCentered*ageCenteredSquared + 
    (1 + ageCentered | ID), # model fails to converge with a random effect for ageCenteredSquared
  data = mydata_long,
  na.action = na.exclude)
```

## Model Summary

```{r}
summary(withinBetweenPersonSleepModel)

print(effectsize::standardize_parameters(
  withinBetweenPersonSleepModel,
  method = "refit"),
  digits = 2)

performance::r2(withinBetweenPersonSleepModel)
```

## Plots

### Between-Person Effect of Sleep

```{r}
#| fig-cap: "Between-Person Association Between Sleep and Depression"

newData <- expand.grid(
  sexFactor = c("female", "male"),
  age = seq(
    from = min(mydata_long$age, na.rm = TRUE),
    to = max(mydata_long$age, na.rm = TRUE),
    length.out = 10000
  ),
  sleepFactor = c("low sleep", "high sleep")
)

newData$ageCentered <- newData$age - min(newData$age)
newData$ageCenteredSquared <- newData$ageCentered ^ 2

newData$sleep_personMean <- NA
newData$sleep_personMean[which(newData$sleepFactor == "low sleep" & newData$sexFactor == "female")] <- mean(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "female")], na.rm = TRUE) - sd(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "female")], na.rm = TRUE)

newData$sleep_personMean[which(newData$sleepFactor == "low sleep" & newData$sexFactor == "male")] <- mean(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "male")], na.rm = TRUE) - sd(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "male")], na.rm = TRUE)

newData$sleep_personMean[which(newData$sleepFactor == "high sleep" & newData$sexFactor == "female")] <- mean(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "female")], na.rm = TRUE) + sd(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "female")], na.rm = TRUE)

newData$sleep_personMean[which(newData$sleepFactor == "high sleep" & newData$sexFactor == "male")] <- mean(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "male")], na.rm = TRUE) + sd(mydata_long$sleep_personMean[which(mydata_long$sexFactor == "male")], na.rm = TRUE)

newData$sleep_personMeanCentered <- 0

newData$modelImpliedDepression <- predict( # predict.merMod
  withinBetweenPersonSleepModel,
  newdata = newData,
  re.form = NA # include no random effects
)

ggplot(
  data = newData,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor,
    linetype = sleepFactor)) +
  geom_line() +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  theme_classic()
```

### Within-Person Effect of Sleep

```{r}
#| fig-cap: "Within-Person Association Between Sleep and Depression"

newData <- expand.grid(
  sexFactor = c("female", "male"),
  age = seq(
    from = min(mydata_long$age, na.rm = TRUE),
    to = max(mydata_long$age, na.rm = TRUE),
    length.out = 10000
  ),
  sleepFactor = c("low sleep", "high sleep")
)

newData$ageCentered <- newData$age - min(newData$age)
newData$ageCenteredSquared <- newData$ageCentered ^ 2

newData$sleep_personMean <- mean(mydata_long$sleep_personMean, na.rm = TRUE)

newData$sleep_personMeanCentered <- NA
newData$sleep_personMeanCentered[which(newData$sleepFactor == "low sleep" & newData$sexFactor == "female")] <- mean(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "female")], na.rm = TRUE) - sd(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "female")], na.rm = TRUE)

newData$sleep_personMeanCentered[which(newData$sleepFactor == "low sleep" & newData$sexFactor == "male")] <- mean(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "male")], na.rm = TRUE) - sd(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "male")], na.rm = TRUE)

newData$sleep_personMeanCentered[which(newData$sleepFactor == "high sleep" & newData$sexFactor == "female")] <- mean(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "female")], na.rm = TRUE) + sd(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "female")], na.rm = TRUE)

newData$sleep_personMeanCentered[which(newData$sleepFactor == "high sleep" & newData$sexFactor == "male")] <- mean(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "male")], na.rm = TRUE) + sd(mydata_long$sleep_personMeanCentered[which(mydata_long$sexFactor == "male")], na.rm = TRUE)

newData$modelImpliedDepression <- predict( # predict.merMod
  withinBetweenPersonSleepModel,
  newdata = newData,
  re.form = NA # include no random effects
)

ggplot(
  data = newData,
  mapping = aes(
    x = age,
    y = modelImpliedDepression,
    color = sexFactor,
    linetype = sleepFactor)) +
  geom_line() +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  theme_classic()
```
