---
title: "Structural Equation Models"
---

For an overview of structural equation modeling, see my chapter on "[Structural Equation Modeling](https://isaactpetersen.github.io/Principles-Psychological-Assessment/structural-equation-modeling.html)" in the following textbook:

Petersen, I. T. (2025). *Principles of psychological assessment: With applied examples in R*. University of Iowa Libraries. <https://isaactpetersen.github.io/Principles-Psychological-Assessment>. <https://doi.org/10.25820/work.007199>

# Preamble

```{r}
#| include: false

options(scipen = 999) # suppress scientific notation
```

## Load Libraries

```{r}
library("lavaan")
library("semTools")
library("lcsm")
library("lavaanPlot")
library("lavaangui")
library("tidyverse")
```

## Load Data

```{r}
mydata_long <- read.csv("./data/mydata_long.csv")

longitudinalMI1 <- read.csv("./data/Bliese-Ployhart-2002-indicators-1.csv")
```

```{r}
#| eval: false

longitudinalMI2 <- read.csv("./data/Bliese-Ployhart-2002-indicators-2.csv")
longitudinalMI3 <- read.csv("./data/Bliese-Ployhart-2002-indicators-3.csv")
longitudinalMI4 <- read.csv("./data/Bliese-Ployhart-2002-indicators-4.csv")
```

# Research Questions {#sec-researchQuestions}

1. What is the form of adolescents' depressive symptom trajectories across ages 14 to 17?
1. Do the trajectories differ between boys and girls?
1. Does how much sleep the adolescent receives predict their depressive symptoms over time?
1. Does the association between sleep and depressive symptoms differ between boys and girls?
1. What is the association between sleep and depressive symptoms at the within-individual level versus between-individual level?
1. Does level of anxiety predict later within-person change in depression (and vice versa)?
1. Does within-person change in anxiety predict later within-person change in depression (and vice versa)?
1. Do our measures show measurement invariance across time?
That is, are our scores able to meaningfully compared on the same metric across time?

# Pre-Model Computation {#sec-premodelComputation}

```{r}
mydata_long$sexFactor <- factor(mydata_long$sex)

mydata_long <- mydata_long |>
  group_by(ID) |>
  mutate(
    # Compute person mean and mean-centered variables
    ses_personMean = mean(ses, na.rm = TRUE),
    ses_personMeanCentered = ses - ses_personMean,
    sleep_personMean = mean(sleep, na.rm = TRUE),
    sleep_personMeanCentered = sleep - sleep_personMean,
    depression_personMean = mean(depression, na.rm = TRUE),
    depression_personMeanCentered = depression - depression_personMean,
    anxiety_personMean = mean(anxiety, na.rm = TRUE),
    anxiety_personMeanCentered = anxiety - anxiety_personMean,
    
    # Compute interaction terms
    sexEffectCodeXses = sexEffectCode * ses,
    sexEffectCodeXsleep = sexEffectCode * sleep,
    sexEffectCodeXdepression = sexEffectCode * depression,
    sexEffectCodeXanxiety = sexEffectCode * anxiety,
    sexEffectCodeXsleep_personMean = sexEffectCode * sleep_personMean,
    sexEffectCodeXsleep_personMeanCentered = sexEffectCode * sleep_personMeanCentered
  ) |>
  ungroup()

# Convert data from long to wide (widen data by timepoint)
mydata_wide <- mydata_long |>
  select(-age) |>
  tidyr::pivot_wider(
    names_from  = timepoint, # widen by timepoint
    values_from = c( # widen time-varying variables
      ses, ses_personMeanCentered, sleep, sleep_personMeanCentered, depression,
      depression_personMeanCentered, anxiety, anxiety_personMeanCentered, 
      sexEffectCodeXses, sexEffectCodeXsleep, sexEffectCodeXdepression, sexEffectCodeXanxiety,
      sexEffectCodeXsleep_personMeanCentered)
  )
```

Variable names for the data in wide form:

```{r}
names(mydata_wide)
```

# Latent Growth Curve Model {#sec-latentGCM}

## Linear Growth Curve Model {#sec-linearGCM}

### Model Syntax

In estimating linear latent growth curves of depression symptoms, this model specifies:

1. latent intercept and slope factors
1. the person's sex (`sexEffectCode`) and the person's average sleep (`sleep_personMean`; and their interaction; `sexEffectCodeXsleep_personMean`) as time-invariant predictors of the latent intercept and slope, and
1. a person's deviations from their average sleep (`sleep_personMeanCentered`; and in interaction with sex; `sexEffectCodeXsleep_personMeanCentered`) as time-varying covariates predicting concurrent depression symptoms at each wave.

```{r}
linearGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  slope =~ 0*depression_1 + 1*depression_2 + 2*depression_3 + 3*depression_4

  # Regression paths
  intercept ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  slope ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  
  # Time-varying covariates
  depression_1 ~ sleep_personMeanCentered_1 + sexEffectCodeXsleep_personMeanCentered_1
  depression_2 ~ sleep_personMeanCentered_2 + sexEffectCodeXsleep_personMeanCentered_2
  depression_3 ~ sleep_personMeanCentered_3 + sexEffectCodeXsleep_personMeanCentered_3
  depression_4 ~ sleep_personMeanCentered_4 + sexEffectCodeXsleep_personMeanCentered_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0*1
  depression_2 ~ 0*1
  depression_3 ~ 0*1
  depression_4 ~ 0*1
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  slope ~ 1
'
```

### Fit Model

Specifying `fixed.x = FALSE` would estimate (i.e., would not consider as fixed) means, variances, and covariances of the observed variables.
Doing so would allow missing information in the predictors to be included as part of full information maximum likelihood (FIML).
However, this gets challenging to use when including both person mean and person-mean-centered variables (due to linear dependencies).
Thus, many of these examples do not specify `fixed.x = FALSE`.

```{r}
linearGCM_fit <- lavaan::sem(
  linearGCM_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE) # estimate means/intercepts
```

### Model Summary

```{r}
summary(
  linearGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  linearGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  linearGCM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(linearGCM_fit)
```

#### Model-Implied Trajectories

```{r}
linearGCM_fs <- as.data.frame(lavaan::lavPredict(linearGCM_fit))
linearGCM_caseidx <- lavaan::lavInspect(linearGCM_fit, "case.idx")
linearGCM_fs$ID <- mydata_wide$ID[linearGCM_caseidx]

factorScores_linear <- left_join(
  mydata_wide,
  linearGCM_fs,
  by = "ID")

get_params <- function(fit) {
  pe <- parameterEstimates(fit)
  
  get <- function(lhs, rhs, op = NULL) {
    if (is.null(op)) {
      # automatically detect the operator
      # priority: =~, ~1, ~
      sub <- pe[pe$lhs == lhs & pe$rhs == rhs, ]
      if (nrow(sub) == 0) stop(paste("Parameter not found:", lhs, rhs))
      out <- sub$est[1]
    } else {
      sub <- pe[pe$lhs == lhs & pe$rhs == rhs & pe$op == op, ]
      if (nrow(sub) == 0) stop(paste("Parameter not found:", lhs, op, rhs))
      out <- sub$est[1]
    }
    return(out)
  }
  
  list(pe = pe, get = get)
}

linearGCM_params <- get_params(linearGCM_fit)

long_linear <- factorScores_linear |>
  pivot_longer(
    cols = matches("depression_|sleep_personMeanCentered_"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)_(\\d)"
  ) |>
  mutate(
    time = as.numeric(time),
    time_num = time - 1,
    age = recode_values(time, 1 ~ 14, 2 ~ 15, 3 ~ 16, 4 ~ 17)
  )

# Compute individuals' model-implied trajectories
long_linear <- long_linear |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ linearGCM_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ linearGCM_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ linearGCM_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ linearGCM_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    
    beta_int = case_when(
      time_num == 0 ~ linearGCM_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ linearGCM_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ linearGCM_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ linearGCM_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    
    yhat =
      intercept +
      slope * time_num +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered
  )

sleep_stats <- mydata_long |>
  group_by(sexFactor, age) |>
  summarize(
    mean_sleep = mean(sleep, na.rm = TRUE),
    sd_sleep = sd(sleep, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(
    time = recode_values(age, 14 ~ 1, 15 ~ 2, 16 ~ 3, 17 ~ 4),
    time_num = time - 1
  )

newData <- expand.grid(
  sexFactor = c("female", "male"),
  time_num = 0:3,
  sleepFactor = c("low sleep", "average sleep", "high sleep")
) |>
  left_join(sleep_stats, by = c("sexFactor", "time_num")) |>
  mutate(
    sleep = case_when(
      sleepFactor == "average sleep" ~ mean_sleep,
      sleepFactor == "low sleep" ~ mean_sleep - sd_sleep,
      sleepFactor == "high sleep" ~ mean_sleep + sd_sleep
    ),
    sexEffectCode = ifelse(sexFactor == "male", 1, -1)
  )

# Compute prototypical trajectories
grand_mean_sleep <- mean(mydata_wide$sleep_personMean, na.rm = TRUE)

newData <- newData |>
  mutate(
    sleep_personMeanCentered = sleep - grand_mean_sleep
  )

newData <- newData |>
  mutate(
    intercept_hat =
    linearGCM_params$get("intercept", "", "~1") +
    linearGCM_params$get("intercept", "sexEffectCode", "~") * sexEffectCode +
    linearGCM_params$get("intercept", "sleep_personMean", "~") * sleep +
    linearGCM_params$get("intercept", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep,
    
  slope_hat =
    linearGCM_params$get("slope", "", "~1") +
    linearGCM_params$get("slope", "sexEffectCode", "~") * sexEffectCode +
    linearGCM_params$get("slope", "sleep_personMean", "~") * sleep +
    linearGCM_params$get("slope", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep
  )

newData <- newData |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ linearGCM_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ linearGCM_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ linearGCM_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ linearGCM_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    
    beta_int = case_when(
      time_num == 0 ~ linearGCM_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ linearGCM_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ linearGCM_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ linearGCM_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    
    yhat =
      intercept_hat +
      slope_hat * time_num +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered
  )
```

##### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor)
) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    se = FALSE
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

ggplot(
  data = long_linear,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID,
    color = sexFactor)
) +
  geom_smooth( # fit a smoothed line to each participant's points; use geom_line() to fit a non-smooth line
    method = "lm",
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = long_linear,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID
  )
) +
  geom_smooth( # fit a smoothed line to each participant's points; use geom_line() to fit a non-smooth line
    method = "lm",
    se = FALSE,
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    data = newData |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = yhat,
      group = sexFactor,
      color = sexFactor
      ),
    method = "lm",
    se = FALSE,
    linewidth = 2
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor,
    linetype = sleepFactor)
) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    se = FALSE
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  guides(
    linetype = guide_legend(
      override.aes = list(color = "black")
    )
  ) +
  theme_classic()
```

## Quadratic Growth Curve Model {#sec-quadraticGCM}

### Model Syntax

In estimating quadratic latent growth curves of depression symptoms, this model specifies:

1. latent intercept and latent linear and quadratic slope factors
1. the person's sex (`sexEffectCode`) and the person's average sleep (`sleep_personMean`; and their interaction; `sexEffectCodeXsleep_personMean`) as time-invariant predictors of the latent intercept and slope, and
1. a person's deviations from their average sleep (`sleep_personMeanCentered`; and in interaction with sex; `sexEffectCodeXsleep_personMeanCentered`) as time-varying covariates predicting concurrent depression symptoms at each wave.

```{r}
quadraticGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  linear =~ 0*depression_1 + 1*depression_2 + 2*depression_3 + 3*depression_4
  quadratic =~ 0*depression_1 + 1*depression_2 + 4*depression_3 + 9*depression_4

  # Regression paths
  intercept ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  linear ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  quadratic ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  
  # Time-varying covariates
  depression_1 ~ sleep_personMeanCentered_1 + sexEffectCodeXsleep_personMeanCentered_1
  depression_2 ~ sleep_personMeanCentered_2 + sexEffectCodeXsleep_personMeanCentered_2
  depression_3 ~ sleep_personMeanCentered_3 + sexEffectCodeXsleep_personMeanCentered_3
  depression_4 ~ sleep_personMeanCentered_4 + sexEffectCodeXsleep_personMeanCentered_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0*1
  depression_2 ~ 0*1
  depression_3 ~ 0*1
  depression_4 ~ 0*1
  
  # Constrain residual variances to be equal (needed to address negative variance)
  depression_1 ~~ vardep*depression_1
  depression_2 ~~ vardep*depression_2
  depression_3 ~~ vardep*depression_3
  depression_4 ~~ vardep*depression_4
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  linear ~ 1
  quadratic ~ 1
'
```

### Fit Model

```{r}
quadraticGCM_fit <- lavaan::sem(
  quadraticGCM_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE) # estimate means/intercepts
```

### Model Summary

```{r}
summary(
  quadraticGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  quadraticGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  quadraticGCM_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(quadraticGCM_fit)
```

#### Model-Implied Trajectories

```{r}
# Factor scores (latent growth factors) for each individual
quadraticGCM_fs <- as.data.frame(lavaan::lavPredict(quadraticGCM_fit))
quadraticGCM_caseidx <- lavaan::lavInspect(quadraticGCM_fit, "case.idx")
quadraticGCM_fs$ID <- mydata_wide$ID[quadraticGCM_caseidx]

factorScores_quad <- left_join(
  mydata_wide,
  quadraticGCM_fs,
  by = "ID"
)

# Function to extract parameter estimates
quad_params <- get_params(quadraticGCM_fit)

# Compute individuals' model-implied trajectories
long_quad <- factorScores_quad |>
  pivot_longer(
    cols = matches("depression_|sleep_personMeanCentered_"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)_(\\d)"
  ) |>
  mutate(
    time = as.numeric(time),
    time_num = time - 1,
    age = recode_values(time, 1 ~ 14, 2 ~ 15, 3 ~ 16, 4 ~ 17)
  ) |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ quad_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ quad_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ quad_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ quad_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    
    beta_int = case_when(
      time_num == 0 ~ quad_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ quad_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ quad_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ quad_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    
    yhat =
      intercept +
      linear * time_num +
      quadratic * time_num^2 +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered
  )

# Compute prototypical trajectories
grand_mean_sleep <- mean(mydata_wide$sleep_personMean, na.rm = TRUE)

newData_quad <- expand.grid(
  sexFactor = c("female", "male"),
  time_num = 0:3,
  sleepFactor = c("low sleep", "average sleep", "high sleep")
) |>
  left_join(sleep_stats, by = c("sexFactor", "time_num")) |>
  mutate(
    sleep = case_when(
      sleepFactor == "average sleep" ~ mean_sleep,
      sleepFactor == "low sleep" ~ mean_sleep - sd_sleep,
      sleepFactor == "high sleep" ~ mean_sleep + sd_sleep
    ),
    sexEffectCode = ifelse(sexFactor == "male", 1, -1),
    sleep_personMeanCentered = sleep - grand_mean_sleep
  ) |>
  mutate(
    intercept_hat =
      quad_params$get("intercept", "", "~1") +
      quad_params$get("intercept", "sexEffectCode", "~") * sexEffectCode +
      quad_params$get("intercept", "sleep_personMean", "~") * sleep +
      quad_params$get("intercept", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep,
    
    linear_hat =
      quad_params$get("linear", "", "~1") +
      quad_params$get("linear", "sexEffectCode", "~") * sexEffectCode +
      quad_params$get("linear", "sleep_personMean", "~") * sleep +
      quad_params$get("linear", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep,
    
    quadratic_hat =
      quad_params$get("quadratic", "", "~1") +
      quad_params$get("quadratic", "sexEffectCode", "~") * sexEffectCode +
      quad_params$get("quadratic", "sleep_personMean", "~") * sleep +
      quad_params$get("quadratic", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep
  ) |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ quad_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ quad_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ quad_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ quad_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    beta_int = case_when(
      time_num == 0 ~ quad_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ quad_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ quad_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ quad_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    yhat =
      intercept_hat +
      linear_hat * time_num +
      quadratic_hat * time_num^2 +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered,
    age = time_num + 14
  )
```

##### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData_quad |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor)
) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

ggplot(
  data = long_quad,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID,
    color = sexFactor)
) +
  geom_smooth( # fit a smoothed line to each participant's points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = long_quad,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID
  )
) +
  geom_smooth( # fit a smoothed line to each participant's points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    data = newData_quad |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = yhat,
      group = sexFactor,
      color = sexFactor
      ),
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    linewidth = 2
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData_quad |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor,
    linetype = sleepFactor)
) +
  geom_smooth( # fit a smoothed line to the model-implied points; use geom_line() to fit a non-smooth line
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  guides(
    linetype = guide_legend(
      override.aes = list(color = "black")
    )
  ) +
  theme_classic()
```

## Latent Basis Growth Curve Model {#sec-latentBasisGCM}

### Model Syntax

In estimating latent basis growth curves of depression symptoms, this model specifies:

1. latent intercept and slope factors, with factor loadings for the slope freely estimated at timepoints 2 and 3 to capture nonlinear patterns of change
1. the person's sex (`sexEffectCode`) and the person's average sleep (`sleep_personMean`; and their interaction; `sexEffectCodeXsleep_personMean`) as time-invariant predictors of the latent intercept and slope, and
1. a person's deviations from their average sleep (`sleep_personMeanCentered`; and in interaction with sex; `sexEffectCodeXsleep_personMeanCentered`) as time-varying covariates predicting concurrent depression symptoms at each wave.

```{r}
latentBasisGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  slope =~ 0*depression_1 + a*depression_2 + b*depression_3 + 1*depression_4 # freely estimate the loadings for t2 and t3

  # Regression paths
  intercept ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  slope ~ sexEffectCode + sleep_personMean + sexEffectCodeXsleep_personMean
  
  # Time-varying covariates
  depression_1 ~ sleep_personMeanCentered_1 + sexEffectCodeXsleep_personMeanCentered_1
  depression_2 ~ sleep_personMeanCentered_2 + sexEffectCodeXsleep_personMeanCentered_2
  depression_3 ~ sleep_personMeanCentered_3 + sexEffectCodeXsleep_personMeanCentered_3
  depression_4 ~ sleep_personMeanCentered_4 + sexEffectCodeXsleep_personMeanCentered_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0*1
  depression_2 ~ 0*1
  depression_3 ~ 0*1
  depression_4 ~ 0*1
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  slope ~ 1
'
```

### Fit Model

```{r}
latentBasisGCM_fit <- lavaan::sem(
  latentBasisGCM_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE) # estimate means/intercepts
```

A model with negative variances is uninterpretable.
We can constrain the variances to be nonnegative using the `bounds = "pos.var"` argument:

```{r}
latentBasisGCM_fit <- lavaan::sem(
  latentBasisGCM_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  bounds = "pos.var") # constrain observed and latent variances to be nonnegative
```

### Model Summary

```{r}
summary(
  latentBasisGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  latentBasisGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  latentBasisGCM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(latentBasisGCM_fit)
```

#### Model-Implied Trajectories

```{r}
# Factor scores for each individual
latentBasisGCM_fs <- as.data.frame(lavaan::lavPredict(latentBasisGCM_fit))
latentBasisGCM_caseidx <- lavaan::lavInspect(latentBasisGCM_fit, "case.idx")
latentBasisGCM_fs$ID <- mydata_wide$ID[latentBasisGCM_caseidx]

factorScores_lb <- left_join(mydata_wide, latentBasisGCM_fs, by = "ID")

# Function to get parameters (reuse your get_params function)
lb_params <- get_params(latentBasisGCM_fit)

# Get estimated factor loadings
lambda2 <- lb_params$get("slope", "depression_2", "=~") # estimated loading 'a'
lambda3 <- lb_params$get("slope", "depression_3", "=~") # estimated loading 'b'
lambda_vec <- c(0, lambda2, lambda3, 1)

# Compute individuals' model-implied trajectories
long_lb <- factorScores_lb |>
  pivot_longer(
    cols = matches("depression_|sleep_personMeanCentered_"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)_(\\d)"
  ) |>
  mutate(
    time = as.numeric(time),
    time_num = time - 1,
    age = recode_values(time, 1 ~ 14, 2 ~ 15, 3 ~ 16, 4 ~ 17),
    lambda = lambda_vec[time] # latent basis loading for each time
  ) |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ lb_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ lb_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ lb_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ lb_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    beta_int = case_when(
      time_num == 0 ~ lb_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ lb_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ lb_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ lb_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    yhat =
      intercept +
      slope * lambda +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered
  )

# Compute prototypical trajectories
grand_mean_sleep <- mean(mydata_wide$sleep_personMean, na.rm = TRUE)

newData_lb <- expand.grid(
  sexFactor = c("female", "male"),
  time_num = 0:3,
  sleepFactor = c("low sleep", "average sleep", "high sleep")
) |>
  left_join(sleep_stats, by = c("sexFactor", "time_num")) |>
  mutate(
    sleep = case_when(
      sleepFactor == "average sleep" ~ mean_sleep,
      sleepFactor == "low sleep" ~ mean_sleep - sd_sleep,
      sleepFactor == "high sleep" ~ mean_sleep + sd_sleep
    ),
    sexEffectCode = ifelse(sexFactor == "male", 1, -1),
    sleep_personMeanCentered = sleep - grand_mean_sleep,
    lambda = lambda_vec[time_num + 1] # latent basis loading for each time
  ) |>
  mutate(
    intercept_hat =
      lb_params$get("intercept", "", "~1") +
      lb_params$get("intercept", "sexEffectCode", "~") * sexEffectCode +
      lb_params$get("intercept", "sleep_personMean", "~") * sleep +
      lb_params$get("intercept", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep,
    slope_hat =
      lb_params$get("slope", "", "~1") +
      lb_params$get("slope", "sexEffectCode", "~") * sexEffectCode +
      lb_params$get("slope", "sleep_personMean", "~") * sleep +
      lb_params$get("slope", "sexEffectCodeXsleep_personMean", "~") * sexEffectCode * sleep
  ) |>
  mutate(
    beta_sleep = case_when(
      time_num == 0 ~ lb_params$get("depression_1", "sleep_personMeanCentered_1"),
      time_num == 1 ~ lb_params$get("depression_2", "sleep_personMeanCentered_2"),
      time_num == 2 ~ lb_params$get("depression_3", "sleep_personMeanCentered_3"),
      time_num == 3 ~ lb_params$get("depression_4", "sleep_personMeanCentered_4")
    ),
    beta_int = case_when(
      time_num == 0 ~ lb_params$get("depression_1", "sexEffectCodeXsleep_personMeanCentered_1"),
      time_num == 1 ~ lb_params$get("depression_2", "sexEffectCodeXsleep_personMeanCentered_2"),
      time_num == 2 ~ lb_params$get("depression_3", "sexEffectCodeXsleep_personMeanCentered_3"),
      time_num == 3 ~ lb_params$get("depression_4", "sexEffectCodeXsleep_personMeanCentered_4")
    ),
    yhat =
      intercept_hat +
      slope_hat * lambda +
      beta_sleep * sleep_personMeanCentered +
      beta_int * sexEffectCode * sleep_personMeanCentered,
    age = time_num + 14
  )
```

##### Prototypical Growth Curve By Sex

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex"

ggplot(
  data = newData_lb |> filter(sleepFactor == "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor)
) +
  geom_line() +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Growth Curves

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms"

ggplot(
  data = long_lb,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID,
    color = sexFactor)
) +
  geom_line(
    alpha = 0.5
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Individuals' Trajectories Overlaid with Prototypical Trajectory By Sex

```{r}
#| fig-cap: "Individuals' Model-Implied Growth Curves of Depression Symptoms Overlaid With Prototypical Growth Curves By Sex"

ggplot(
  data = long_lb,
  mapping = aes(
    x = age,
    y = yhat,
    group = ID
  )
) +
  geom_line( # individuals' trajectories
    linewidth = 0.5,
    color = "gray",
    alpha = 0.30
  ) +
  geom_line( # prototypical trajectories
    data = newData_lb |> filter(sleepFactor == "average sleep"),
    mapping = aes(
      x = age,
      y = yhat,
      group = sexFactor,
      color = sexFactor
      ),
    linewidth = 2
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex"
  ) +
  theme_classic()
```

##### Prototypical Trajectory By Sex and Sleep

```{r}
#| fig-cap: "Prototypical Model-Implied Growth Curves of Depression Symptoms By Sex and Sleep"

ggplot(
  data = newData_lb |> filter(sleepFactor != "average sleep"),
  mapping = aes(
    x = age,
    y = yhat,
    color = sexFactor,
    linetype = sleepFactor)
) +
  geom_line() +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms",
    color = "Sex",
    linetype = "Sleep"
  ) +
  guides(
    linetype = guide_legend(
      override.aes = list(color = "black")
    )
  ) +
  theme_classic()
```

# Compare Models {#sec-compareModels}

```{r}
anova(
  linearGCM_fit,
  quadraticGCM_fit
)

anova(
  linearGCM_fit,
  latentBasisGCM_fit
)

anova(
  latentBasisGCM_fit,
  quadraticGCM_fit
)
```

# Latent Change Score Model {#sec-lcsm}

## Model Syntax

This model examines reciprocal (i.e., bidirectional), dynamic (i.e., changing) relations between latent change in depression and latent change in anxiety.
Each construct is modeled with a latent intercept (constant change factor), a proportional change factor (latent true score predicting its own change), and autoregressive effects of prior change scores.
The cross-domain coupling parameters evaluate whether latent change in depression predicts subsequent latent change in anxiety, and vice versa.

```{r}
bivariateLCSM_syntax <- lcsm::specify_bi_lcsm(
  timepoints = 4,
  var_x = "depression_",
  model_x = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  var_y = "anxiety_",
  model_y = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  coupling = list(
    delta_lag_xy = TRUE, # level of x predicting change of y
    delta_lag_yx = TRUE, # level of y predicting change of x
    xi_lag_xy = TRUE,    # change of x predicting change of y
    xi_lag_yx = TRUE),   # change of y predicting change of x
  change_letter_x = "g",
  change_letter_y = "j")

cat(bivariateLCSM_syntax)
```

## Fit Model

```{r}
bivariateLCSM_fit <- lcsm::fit_bi_lcsm(
  data = mydata_wide,
  var_x = c("depression_1","depression_2","depression_3","depression_4"),
  var_y = c("anxiety_1","anxiety_2","anxiety_3","anxiety_4"),
  model_x = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  model_y = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  coupling = list(
    delta_lag_xy = TRUE, # level of x predicting change of y
    delta_lag_yx = TRUE, # level of y predicting change of x
    xi_lag_xy = TRUE,    # change of x predicting change of y
    xi_lag_yx = TRUE),   # change of y predicting change of x
  fixed.x = FALSE # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
  )
```

## Model Summary

```{r}
summary(
  bivariateLCSM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

lcsm::extract_param(bivariateLCSM_fit)
```

## Modification Indices

```{r}
modificationindices(
  bivariateLCSM_fit,
  sort. = TRUE)
```

## Plots

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  bivariateLCSM_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(bivariateLCSM_fit)
```

```{r}
#| eval: false

lcsm::plot_lcsm(
  lavaan_object = bivariateLCSM_fit,
  lcsm = "bivariate",
  lavaan_syntax = bivariateLCSM_syntax,
  edge.label.cex = .9,
  lcsm_colours = TRUE)
```

### Model-Implied Trajectories

To plot model-implied trajectories, see this resource:

<https://thechangelab.stanford.edu/tutorials/growth-modeling/latent-change-score-modeling-lavaan/> (archived at <https://perma.cc/A2KZ-QPD9>)

```{r}
modelPredictedValues_wide <- as.data.frame(lavPredict(bivariateLCSM_fit, type = "yhat"))

modelPredictedValues_wide$ID <- mydata_wide$ID

modelPredictedValues_long <- modelPredictedValues_wide |>
  pivot_longer(
    cols = -ID,
    names_to = c("construct", "time"),
    names_pattern = "([xy])(\\d)",
    values_to = "score"
  ) |>
  mutate(
    time = as.integer(time),
    age = recode_values(time, 1 ~ 14, 2 ~ 15, 3 ~ 16, 4 ~ 17),
    construct = recode_values(construct, "x" ~ "depression", "y" ~ "anxiety")) |>
  pivot_wider(
    names_from = construct,
    values_from = score
  )
```

```{r}
#| fig-cap: "Model-Implied Trajectories of Depression Symptoms"

ggplot(
  data = modelPredictedValues_long,
  mapping = aes(
    x = age,
    y = depression,
    group = ID)) +
  geom_line(
    alpha = 0.25
  ) +
  labs(
    x = "Age (years)",
    y = "Depression Symptoms"
  ) +
  theme_classic()
```

```{r}
#| fig-cap: "Model-Implied Trajectories of Anxiety Symptoms"

ggplot(
  data = modelPredictedValues_long,
  mapping = aes(
    x = time,
    y = anxiety,
    group = ID)) +
  geom_line(
    alpha = 0.25
  ) +
  labs(
    x = "Age (years)",
    y = "Anxiety Symptoms"
  ) +
  theme_classic()
```

# Cross-Lagged Panel Model {#sec-clpm}

## Model Syntax

This cross-lagged panel model examines whether anxiety predicts later depression (i.e., cross-lagged paths), controlling for prior levels of depression (i.e., autoregressive effects), and simultaneously examines the reverse association (i.e., depression predicting later anxiety, controlling for prior levels of anxiety).
Concurrent covariances between depression and anxiety at each timepoint account for shared variance at the same occasion.

```{r}
clpm_syntax <- '
  # Autoregressive Paths
  depression_4 ~ depression_3
  depression_3 ~ depression_2
  depression_2 ~ depression_1
  
  anxiety_4 ~ anxiety_3
  anxiety_3 ~ anxiety_2
  anxiety_2 ~ anxiety_1
  
  # Concurrent Covariances
  depression_1 ~~ anxiety_1
  depression_2 ~~ anxiety_2
  depression_3 ~~ anxiety_3
  depression_4 ~~ anxiety_4
  
  # Cross-Lagged Paths
  depression_4 ~ anxiety_3
  depression_3 ~ anxiety_2
  depression_2 ~ anxiety_1
  
  anxiety_4 ~ depression_3
  anxiety_3 ~ depression_2
  anxiety_2 ~ depression_1
'
```

## Fit Model

```{r}
clpm_fit <- lavaan::sem(
  clpm_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

## Model Summary

```{r}
summary(
  clpm_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
modificationindices(
  clpm_fit,
  sort. = TRUE)
```

## Plots

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  clpm_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(clpm_fit)
```

# Random-Intercepts Cross-Lagged Panel Model {#sec-riclpm}

Adapted from:

- Mulder & Hamaker (2021): <https://doi.org/10.1080/10705511.2020.1784738>
  - <https://jeroendmulder.github.io/RI-CLPM/lavaan.html> (archived at <https://perma.cc/2K6A-WUJQ>)
- Mund & Nestler (2017): <https://osf.io/a4dhk>

## Model Syntax

This random-intercepts cross-lagged panel model examines whether anxiety predicts later depression (i.e., cross-lagged paths), controlling for prior levels of depression (i.e., autoregressive effects) and controlling for a person's general latent level of depression, and simultaneously and simultaneously examines the reverse association (i.e., depression predicting later anxiety, controlling for prior levels of anxiety and general latent levels of anxiety).
Concurrent covariances between depression and anxiety at each timepoint account for shared variance at the same occasion.

```{r}
riclpm_syntax <- '
  # Random Intercepts
  depression =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  anxiety =~ 1*anxiety_1 + 1*anxiety_2 + 1*anxiety_3 + 1*anxiety_4
  
  # Create Within-Person Centered Variables
  wdep1 =~ 1*depression_1
  wdep2 =~ 1*depression_2
  wdep3 =~ 1*depression_3
  wdep4 =~ 1*depression_4
  
  wanx1 =~ 1*anxiety_1
  wanx2 =~ 1*anxiety_2
  wanx3 =~ 1*anxiety_3
  wanx4 =~ 1*anxiety_4
  
  # Autoregressive Paths
  wdep4 ~ wdep3
  wdep3 ~ wdep2
  wdep2 ~ wdep1
  
  wanx4 ~ wanx3
  wanx3 ~ wanx2
  wanx2 ~ wanx1
  
  # Concurrent Covariances
  wdep1 ~~ wanx1
  wdep2 ~~ wanx2
  wdep3 ~~ wanx3
  wdep4 ~~ wanx4
  
  # Cross-Lagged Paths
  wdep4 ~ wanx3
  wdep3 ~ wanx2
  wdep2 ~ wanx1
  
  wanx4 ~ wdep3
  wanx3 ~ wdep2
  wanx2 ~ wdep1
  
  # Variance and Covariance of Random Intercepts
  depression ~~ depression
  anxiety ~~ anxiety
  depression ~~ anxiety
  
  # Variances of Within-Person Centered Variables
  wdep1 ~~ wdep1
  wdep2 ~~ wdep2
  wdep3 ~~ wdep3
  wdep4 ~~ wdep4
  
  wanx1 ~~ wanx1
  wanx2 ~~ wanx2
  wanx3 ~~ wanx3
  wanx4 ~~ wanx4
  
  # Fix Error Variances of Observed Variables to Zero
  depression_1 ~~ 0*depression_1
  depression_2 ~~ 0*depression_2
  depression_3 ~~ 0*depression_3
  depression_4 ~~ 0*depression_4
  
  anxiety_1 ~~ 0*anxiety_1
  anxiety_2 ~~ 0*anxiety_2
  anxiety_3 ~~ 0*anxiety_3
  anxiety_4 ~~ 0*anxiety_4
  
  # Fix the Covariances Between the Random Intercepts and the Latents at T1 to Zero
  wdep1 ~~ 0*depression
  wdep1 ~~ 0*anxiety
  
  wanx1 ~~ 0*depression
  wanx1 ~~ 0*anxiety
  
  # Estimate Observed Intercepts
  depression_1 ~ 1
  depression_2 ~ 1
  depression_3 ~ 1
  depression_4 ~ 1
  
  anxiety_1 ~ 1
  anxiety_2 ~ 1
  anxiety_3 ~ 1
  anxiety_4 ~ 1
  
  # Fix the Means of the Latents to Zero
  wdep1 ~ 0*1
  wdep2 ~ 0*1
  wdep3 ~ 0*1
  wdep4 ~ 0*1
  
  wanx1 ~ 0*1
  wanx2 ~ 0*1
  wanx3 ~ 0*1
  wanx4 ~ 0*1
  
  depression ~ 0*1
  anxiety ~ 0*1
'
```

## Fit Model

```{r}
#| warning: true

riclpm_fit <- lavaan::sem(
  riclpm_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  fixed.x = FALSE, # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
  bounds = "standard" # constrains observed variances, latent variances, and factor loadings to be within computed lower and upper bounds
)
```

Modify the syntax to address the model fit errors (don't estimate concurrent covariances):

```{r}
riclpm2_syntax <- '
  # Random Intercepts
  depression =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  anxiety =~ 1*anxiety_1 + 1*anxiety_2 + 1*anxiety_3 + 1*anxiety_4
  
  # Create Within-Person Centered Variables
  wdep1 =~ 1*depression_1
  wdep2 =~ 1*depression_2
  wdep3 =~ 1*depression_3
  wdep4 =~ 1*depression_4
  
  wanx1 =~ 1*anxiety_1
  wanx2 =~ 1*anxiety_2
  wanx3 =~ 1*anxiety_3
  wanx4 =~ 1*anxiety_4
  
  # Autoregressive Paths
  wdep4 ~ wdep3
  wdep3 ~ wdep2
  wdep2 ~ wdep1
  
  wanx4 ~ wanx3
  wanx3 ~ wanx2
  wanx2 ~ wanx1
  
  # Concurrent Covariances
  #wdep1 ~~ wanx1
  #wdep2 ~~ wanx2
  #wdep3 ~~ wanx3
  #wdep4 ~~ wanx4
  
  # Cross-Lagged Paths
  wdep4 ~ wanx3
  wdep3 ~ wanx2
  wdep2 ~ wanx1
  
  wanx4 ~ wdep3
  wanx3 ~ wdep2
  wanx2 ~ wdep1
  
  # Variance and Covariance of Random Intercepts
  depression ~~ depression
  anxiety ~~ anxiety
  depression ~~ anxiety
  
  # Variances of Within-Person Centered Variables
  wdep1 ~~ wdep1
  wdep2 ~~ wdep2
  wdep3 ~~ wdep3
  wdep4 ~~ wdep4
  
  wanx1 ~~ wanx1
  wanx2 ~~ wanx2
  wanx3 ~~ wanx3
  wanx4 ~~ wanx4
  
  # Fix Error Variances of Observed Variables to Zero
  depression_1 ~~ 0*depression_1
  depression_2 ~~ 0*depression_2
  depression_3 ~~ 0*depression_3
  depression_4 ~~ 0*depression_4
  
  anxiety_1 ~~ 0*anxiety_1
  anxiety_2 ~~ 0*anxiety_2
  anxiety_3 ~~ 0*anxiety_3
  anxiety_4 ~~ 0*anxiety_4
  
  # Fix the Covariances Between the Random Intercepts and the Latents at T1 to Zero
  wdep1 ~~ 0*depression
  wdep1 ~~ 0*anxiety
  
  wanx1 ~~ 0*depression
  wanx1 ~~ 0*anxiety
  
  # Estimate Observed Intercepts
  depression_1 ~ 1
  depression_2 ~ 1
  depression_3 ~ 1
  depression_4 ~ 1
  
  anxiety_1 ~ 1
  anxiety_2 ~ 1
  anxiety_3 ~ 1
  anxiety_4 ~ 1
  
  # Fix the Means of the Latents to Zero
  wdep1 ~ 0*1
  wdep2 ~ 0*1
  wdep3 ~ 0*1
  wdep4 ~ 0*1
  
  wanx1 ~ 0*1
  wanx2 ~ 0*1
  wanx3 ~ 0*1
  wanx4 ~ 0*1
  
  depression ~ 0*1
  anxiety ~ 0*1
'

riclpm2_fit <- lavaan::sem(
  riclpm2_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  fixed.x = FALSE, # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
  bounds = "standard" # constrains observed variances, latent variances, and factor loadings to be within computed lower and upper bounds
)
```

## Model Summary

```{r}
summary(
  riclpm2_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
modificationindices(
  riclpm2_fit,
  sort. = TRUE)
```

## Path Diagram

```{r}
lavaanPlot::lavaanPlot(
  riclpm2_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(riclpm2_fit)
```

# Multilevel SEM {#sec-multilevelSEM}

## Model Syntax

This multilevel structural equation model disaggregates within- and between-person effects to examine associations with depression.
At the within-person level, we examine deviations in sleep, anxiety, and SES from a person's usual (average) levels in predicting concurrent deviations in depression, allowing the within-person association between sleep and depression to differ between boys and girls.
At the between-person level, we examine person-level averages across time of sleep, anxiety, and SES, along with sex, in predicting overall differences in depression across individuals, allowing the between-person association between sleep and depression to differ between boys and girls.

```{r}
multilevelSEM_syntax <- '
    level: 1
        depression ~ sleep_personMeanCentered + anxiety_personMeanCentered + ses_personMeanCentered + sexEffectCodeXsleep_personMeanCentered
    level: 2
        depression ~ sleep_personMean + anxiety_personMean + ses_personMean + sexEffectCode + sexEffectCodeXsleep_personMean
'
```

## Fit Model

```{r}
multilevelSEM_fit <- lavaan::sem(
  multilevelSEM_syntax,
  data = mydata_long,
  cluster = "ID",
  missing = "ML", # FIML
  estimator = "MLR")
```

## Model Summary

```{r}
summary(
  multilevelSEM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
modificationindices(
  multilevelSEM_fit,
  sort. = TRUE)
```

## Path Diagram

```{r}
lavaanPlot::lavaanPlot(
  multilevelSEM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(multilevelSEM_fit)
```

# Longitudinal Measurement Invariance {#sec-longitudinalMI}

We will follow the approach suggested by Widaman et al. (2010).

Widaman, K. F., Ferrer, E., & Conger, R. D. (2010). Factorial invariance within longitudinal structural equation models: Measuring the same construct across time. *Child Development Perspectives*, *4*(1), 10–18. <https://doi.org/10.1111/j.1750-8606.2009.00110.x>

For the longitudinal measurement invariance models, we will use simulated data from <https://mycourses.aalto.fi/mod/assign/view.php?id=1203047> (archived at <https://perma.cc/QTL2-ZHX2>).
These data were simulated and adapted from data from Bliese and Ployhart (2002):

Bliese, P. D., & Ployhart, R. E. (2002). Growth modeling using random coefficient models: Model building, testing, and illustrations. *Organizational Research Methods*, *5*(4), 362–387. <https://doi.org/10.1177/109442802237116>

As noted [here](https://mycourses.aalto.fi/mod/assign/view.php?id=1203047) (archived at <https://perma.cc/QTL2-ZHX2>), "The data are in wide format where the variables are coded as [VARIABLE NAME][time index][indicator index]."
The data include three indicators for each of three constructs: 1) job satisfaction (`JOBSAT`), 2) organizational commitment (`COMMIT`), and 3) readiness to perform (`READY`).

Questions:

1. Which level of longitudinal measurement invariance holds for the `Bliese-Ployhart-2002-indicators-1.csv` dataset (saved in object `longitudinalMI1`)?
1. Repeat the tests of longitudinal measurement invariance for the remaining datasets: `Bliese-Ployhart-2000-indicators-2.csv`, `Bliese-Ployhart-2000-indicators-3.csv`, and `Bliese-Ployhart-2000-indicators-4.csv`.
Which level of longitudinal measurement invariance holds for each dataset?

The answers to these questions are [here](answers.qmd).

## Configural Invariance

A configural invariance model has the same number of latent factors across time, with the indicators loading onto the same latent factor(s) across time.
Model fit indices can then be examined to evaluate how tenable it is that (a) there are the same number of latent factors across time and that (b) the indicators load onto the same latent factor(s) across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time

### Model Syntax

```{r}
configuralInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + JOBSAT12 + JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + JOBSAT22 + JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + JOBSAT32 + JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + READY12 + READY13
  ready_2 =~ NA*loadr1*READY21 + READY22 + READY23
  ready_3 =~ NA*loadr1*READY31 + READY32 + READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + COMMIT12 + COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + COMMIT22 + COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + COMMIT32 + COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
'
```

### Fit Model

```{r}
configuralInvariance_fit <- cfa(
  configuralInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

### Model Summary

```{r}
summary(
  configuralInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  configuralInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  configuralInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(configuralInvariance_fit)
```

## Configural Invariance With Correlated Residuals Within-Indicator Across Time

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. **Allow within-indicator residuals to covary across time**

### Model Syntax

```{r}
configuralInvarianceCorrelatedResiduals_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + JOBSAT12 + JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + JOBSAT22 + JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + JOBSAT32 + JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + READY12 + READY13
  ready_2 =~ NA*loadr1*READY21 + READY22 + READY23
  ready_3 =~ NA*loadr1*READY31 + READY32 + READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + COMMIT12 + COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + COMMIT22 + COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + COMMIT32 + COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
configuralInvarianceCorrelatedResiduals_fit <- cfa(
  configuralInvarianceCorrelatedResiduals_syntax,
  data = longitudinalMI1,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

### Model Summary

```{r}
summary(
  configuralInvarianceCorrelatedResiduals_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  configuralInvarianceCorrelatedResiduals_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  configuralInvarianceCorrelatedResiduals_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(configuralInvarianceCorrelatedResiduals_fit)
```

### Compare Models

```{r}
anova(
  configuralInvariance_fit,
  configuralInvarianceCorrelatedResiduals_fit
)
```

## Metric ("Weak Factorial") Invariance

A metric invariance model evaluates whether the items' factor loadings are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. Allow within-indicator residuals to covary across time
1. **For each indicator, constrain its factor loading to be the same across time**

### Model Syntax

```{r}
metricInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
metricInvariance_fit <- cfa(
  metricInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

### Model Summary

```{r}
summary(
  metricInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  metricInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  metricInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(metricInvariance_fit)
```

### Compare Models

```{r}
anova(
  configuralInvarianceCorrelatedResiduals_fit,
  metricInvariance_fit
)
```

## Scalar ("Strong Factorial") Invariance

A scalar invariance model evaluates whether the items' intercepts are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For each indicator, constrain its factor loading to be the same across time
1. **For each indicator, constrain its intercept to be the same across time**

### Model Syntax

```{r}
scalarInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicators Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  JOBSAT12 ~ intj2*1
  JOBSAT22 ~ intj2*1
  JOBSAT32 ~ intj2*1
  JOBSAT13 ~ intj3*1
  JOBSAT23 ~ intj3*1
  JOBSAT33 ~ intj3*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  READY12 ~ intr2*1
  READY22 ~ intr2*1
  READY32 ~ intr2*1
  READY13 ~ intr3*1
  READY23 ~ intr3*1
  READY33 ~ intr3*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  COMMIT12 ~ intc2*1
  COMMIT22 ~ intc2*1
  COMMIT32 ~ intc2*1
  COMMIT13 ~ intc3*1
  COMMIT23 ~ intc3*1
  COMMIT33 ~ intc3*1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
scalarInvariance_fit <- cfa(
  scalarInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

### Model Summary

```{r}
summary(
  scalarInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  scalarInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  scalarInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(scalarInvariance_fit)
```

### Compare Models

```{r}
anova(
  metricInvariance_fit,
  scalarInvariance_fit
)
```

## Residual ("Strict Factorial") Invariance

A residual invariance model evaluates whether the items' residual variances are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For each indicator, constrain its factor loading to be the same across time
1. For each indicator, constrain its intercept to be the same across time
1. **For each indicator, constrain its residual variance to be the same across time**

### Model Syntax

```{r}
residualInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Constrain Intercepts of Indicators Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  JOBSAT12 ~ intj2*1
  JOBSAT22 ~ intj2*1
  JOBSAT32 ~ intj2*1
  JOBSAT13 ~ intj3*1
  JOBSAT23 ~ intj3*1
  JOBSAT33 ~ intj3*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  READY12 ~ intr2*1
  READY22 ~ intr2*1
  READY32 ~ intr2*1
  READY13 ~ intr3*1
  READY23 ~ intr3*1
  READY33 ~ intr3*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  COMMIT12 ~ intc2*1
  COMMIT22 ~ intc2*1
  COMMIT32 ~ intc2*1
  COMMIT13 ~ intc3*1
  COMMIT23 ~ intc3*1
  COMMIT33 ~ intc3*1
  
  # Constrain Residual Variances of Indicators Across Time
  JOBSAT11 ~~ resj1*JOBSAT11
  JOBSAT21 ~~ resj1*JOBSAT21
  JOBSAT31 ~~ resj1*JOBSAT31
  JOBSAT12 ~~ resj2*JOBSAT12
  JOBSAT22 ~~ resj2*JOBSAT22
  JOBSAT32 ~~ resj2*JOBSAT32
  JOBSAT13 ~~ resj3*JOBSAT13
  JOBSAT23 ~~ resj3*JOBSAT23
  JOBSAT33 ~~ resj3*JOBSAT33
  
  READY11 ~~ resr1*READY11
  READY21 ~~ resr1*READY21
  READY31 ~~ resr1*READY31
  READY12 ~~ resr2*READY12
  READY22 ~~ resr2*READY22
  READY32 ~~ resr2*READY32
  READY13 ~~ resr3*READY13
  READY23 ~~ resr3*READY23
  READY33 ~~ resr3*READY33
  
  COMMIT11 ~~ resc1*COMMIT11
  COMMIT21 ~~ resc1*COMMIT21
  COMMIT31 ~~ resc1*COMMIT31
  COMMIT12 ~~ resc2*COMMIT12
  COMMIT22 ~~ resc2*COMMIT22
  COMMIT32 ~~ resc2*COMMIT32
  COMMIT13 ~~ resc3*COMMIT13
  COMMIT23 ~~ resc3*COMMIT23
  COMMIT33 ~~ resc3*COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
residualInvariance_fit <- cfa(
  residualInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE) # estimate means, variances, and covariances of the observed variables (to include predictors as part of FIML)
```

### Model Summary

```{r}
summary(
  residualInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  residualInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  residualInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

To create an interactive path diagram, you can use the following syntax:

```{r}
#| eval: false

lavaangui::plot_lavaan(residualInvariance_fit)
```

### Compare Models

```{r}
anova(
  scalarInvariance_fit,
  residualInvariance_fit
)

round(cbind(
  configural = inspect(configuralInvarianceCorrelatedResiduals_fit, "fit.measures"), 
  weak = inspect(metricInvariance_fit, "fit.measures"),
  strong = inspect(scalarInvariance_fit, "fit.measures"),
  strict = inspect(residualInvariance_fit, "fit.measures")), 3)
```
